Supervised learning approaches to classifying morphological viariation: picking amongst taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})

### 2016-05-06 doc/turtles_statistical_learning.tex
fixed some of the labels for the figures; need to make sure schemes are correct because confusion

\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
%\title{Supervised learning approaches to quantifying cryptic diversity: picking amongst taxonomic hypotheses of Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[4]{Bryan Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Committee on Evolutionary Biology, University of Chicago}
\affil[2]{Integrative Research Center, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
\affil[4]{Section of Research and Collections, North Carolina Museum of Sciences}

@@

% auc of roc as performance measure
%   relationship between false positive rate and true positive rate
%   varies between 0.5 and 1 (random -- perfect)
%   better performance than accuracy when groups unbalanced
In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}. The area under the ROC curce (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classificationrates, respectively \citep{Hastie2009}. Both ROC and AUC are preferrable to simple classificaiton accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}. The standard ROC and AUC calculations are defined only for binary classifications. To generalize this approach for mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.

@@

% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)
The supervised learning methods used here were multinomial logistic regression (MLR), linear discrminiate analysis (LDA), penalized discriminate analysis (PDA), single-hidden-layer neutral network (NN), and random forests (RF). Each of these methods makes different assumptions, treat data differently, and can produce very different classification results \citep{Hastie2009}. The common assumption of all of these methods is that the predictors or features are independent and/or have additive effects on prediction \citep{Hastie2009}.

>>>>

%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
\title{Supervised learning approaches to classifying morphological viariation: picking amongst taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[4]{Bryan Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Committee on Evolutionary Biology, University of Chicago}
\affil[2]{Integrative Research Center, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
\affil[4]{Section of Research and Collections, North Carolina Museum of Sciences}

@@

% auc of roc as performance measure
%   relationship between false positive rate and true positive rate
%   varies between 0.5 and 1 (random -- perfect)
%   better performance than accuracy when groups unbalanced
In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \citep{Hastie2009}. The area under the ROC curce (AUC) is then the derived estimate of the model performance; AUC ranges from 0.5 to 1 which corresponding to performance similar to random guesses and perfect classificationrates, respectively \citep{Hastie2009}. Both ROC and AUC are preferrable to simple classificaiton accuracy when class membership is unbalanced, as it is in these analyses \citep{Hastie2009}. The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our seven species and \textit{Emys} complex datasets. To generalize this approach for situations with mulitple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \citep{Hand2001}.

@@

% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)
Instead of relying on a single supervised learning method, we chose to use an ensemble of mulitple approaches so that the congruence bewteen the them could be used as a means of ``support'' for one conclusion or another. The supervised learning methods used here are described in Table \ref{tab:methods}. Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}.

\begin{table}
  \centering
  \begin{tabular}{l l l l}
    \hline
    Method name & abbreviation & R package & citation \\
    \hline
    multinomial logistic regression & MLR & nnet & citation \\
    linear discriminate analysis & LDA & MASS & citation \\
    penalized discrminiate analysis & PDA & mda & citation \\
    single-hidden-layer neural network & NN & nnet & citation \\
    random forests & RF & randomForest & citation \\
    \hline
  \end{tabular}
  \caption{table of the methods}
  \label{tab:methods}
\end{table}
